Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-11-23 analysis/paper/paper.Rmd
added more about mark-recapture assumptions and more description of the various estimators we used.

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)

@@

```{r schnabel-estimator, eval = T}
# using Schnabel and Schumacher-Eschmeyer estimators
N_mods = cmr_df %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species, date) %>%
  summarise(n = n(),
            m = sum(capture_type == "Recapture", na.rm = T),
            u = sum(capture_type == "Mark" | is.na(capture_type)),
            R = sum(fish_status == "Live", na.rm = T),
            .groups = "drop") %>%
  group_by(event_name, species) %>%
  nest() %>%
  crossing(model = c("Schnabel",
                     "SchumacherEschmeyer")) %>%
  mutate(mr_model = map2(data,
                         model,
                         .f = function(x, y) {
                           with(x,
                                mrClosed(n = n,
                                         m = m,
                                         R = R,
                                         method = y,
                                         chapman.mod = TRUE))
                         }),
         N = map_dbl(mr_model,
                     .f = summary),
         CI = map(mr_model,
                  .f = confint),
         Lci = map_dbl(CI,
                       .f = function(x) x[1]), 
         Uci = map_dbl(CI,
                       .f = function(x) x[2])) %>%
  ungroup() %>%
  select(-CI)

@@

@@

The Lincoln-Petersen estimator can be biased with small samples, so we also investigated the Chapman-modified Lincoln-Petersen estimator which is shown below.
$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$
The Schnabel estimator is shown below, where the $M$, $n$ and $m$ are indexed by the sampling occasion, $i$. In our example, the sampling occasions are defined as each day of sampling. This estimator does not have an associate standard error, but 95% confidence intervals can be calculated.
$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_i}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
There is another estimator for this type of multiple census surveys, called the Schumacher-Eschmeyer estimator, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. It is shown below.

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$
Additionally, we calculated the proportional stock density (PSD) of Northern Pikeminnow with 300 mm total length (TL) for stock and 400 mm TL for quality.
$$
PSD_{i} = 100 * \frac{FQ_{i}}{FS_{i}}
$$
where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$.

@@

```{r shoup-temp-data}
shoup_temp_df = read_tsv(here("analysis/data/raw_data/USGS_13307000_Salmon_River_Nr_Shoup.txt"),
                   comment = "#") %>%
  rename(discharge_cfs = `47768_00060`,
         discharge_cfs_cd = `47768_00060_cd`,
         gage_ht_ft = `47769_00065`,
         gage_ht_ft_cd = `47769_00065_cd`,
         water_temp_c = `47773_00010`,
         water_temp_c_cd = `47773_00010_cd`)

```{r read-bio-outputs}
# full year, no growth
full_yr_no_growth = read_csv(here("analysis/data/raw_data/2020 Bioenergetics 1 year.csv")) %>%
  clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  mutate(cum_cons_fish_g = cumsum(cons_fish_g)) %>%
  mutate(month = month(date))

# full year, 10% growth
full_yr_10p_growth = read_csv(here("analysis/data/raw_data/2020 Bioenergetics 1 year 10% growth.csv")) %>%
  clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  mutate(cum_cons_fish_g = cumsum(cons_fish_g)) %>%
  mutate(month = month(date))

# an alternative model
alternative_mod = read_csv(here("analysis/data/raw_data/Bioenergetics 32 fish 68 inverts.csv")) %>%
  clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  mutate(cum_cons_fish_g = cumsum(cons_fish_g)) %>%
  mutate(month = month(date))

# Fall emigration period (9/1 - 11/17), no growth
fall_only_no_growth = read_csv(here("analysis/data/raw_data/78_day_bioenergetics.csv")) %>%
  clean_names() %>%
  mutate(date = as.Date(date, format = "%m/%d/%y")) %>%
  mutate(cum_cons_fish_g = cumsum(cons_fish_g)) %>%
  mutate(month = month(date))

@@

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer")) %>%
  arrange(event_name, model)

@@

During 2019, 660 Northern Pikeminnow were gastric lavaged. Of those fish, we observed 603 individuals with empty stomachs, and 57 individuals with stomach contents. Twelve of the samples contained fish remnants including a single Mountain Whitefish and two Redside Shiners. During the 2020 season, 805 Northern Pikeminnow were gastric lavaged comprised of 613 individuals with empty stomachs and 188 individuals with stomach contents. Twenty-three individuals had fish remnants in their stomach contents including ten with unidentifiable remnants, six Redside Shiners, four suckers, two sculpin, and one juvenile Chinook Salmon. In 2021, we collected stomach contents from 105 Northern Pikeminnow with 98 containing non-fish prey items and observed seven containing fish remnants.

@@

Over the fall time period a single Northern Pikeminnow is modeled to eat 61 g of fish. For the estimated population of ***xx,xxx***, then ***xxx,xxx*** g of fish are being eaten each fall during the migration of juvenile Chinook Salmon. Therefore, Northern Pikeminnow in the Deadwater Slough can potentially consume ***xx,xxx*** juvenile Chinook Salmon as a discrete population.

Over an entire year, we estimate that a single Northern Pikeminnow will eat 283 g of fish and 613 g of other food types.

@@

The majority of fish collected during this study received gastric lavage, including some non-predatory species. To validate the efficacy of this method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. Our results support previous findings that gastric lavage effectively removes stomach contents (@Lott2020). Additionally, we found that the fish captured in net-traps had a similar proportion of stomach content samples. of those, 1,214 (76.2%) were completely empty, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish remnants. We were successfully able to identify Redside Shiners, a Largescale Sucker, sculpin, Mountain Whitefish, and one juvenile Chinook Salmon in the stomach contents. However, decomposition from digestion rendered most stomach contents unidentifiable.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile salmon in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the associated riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using the Deadwater Slough that include the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the initial sampling event in 2019, a two-person crew walked the entire reach and surrounding and upstream areas scanning for passive integrated transponder (PIT) tags. During that informal survey, nine PIT tags were recovered near active bird nests or in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (3), and steelhead (3). Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake River and Columbia River, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon from the many piscivorous birds known to use the site. Future estimates of predation would benefit from consideration of the contribution of piscivorous avian predators.

@@

We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces annual adult returns by ***xxx - x,xxx*** (perhaps express as percent, too) to upriver populations. Presumably, that reduction in adult returns impacts both the ESA-listed natural populations in the Upper Salmon River MPG and two hatchery populations in the Upper Salmon, Pahsimeroi and Sawtooth hatcheries, which provide for recrational fishing opportunities. Consequently, reducing juvenile Chinook Salmon predation mortalities at Deadwater Slough can potentially benefit multiple upriver natural and hatchery populations upriver. Because the Deadwater Slough is part of the migratory pathway for emigrating salmonids, including for multiple species and populations, the impact of that benefit would likely be higher than tributary habitat rehabilitation actions, which typically benefit a single population. Moreover, the deepened, slack water conditions that favor northern pikeminnow at Deadwater Slough are indirectly the result of manmade activities i.e., the failure of a manmade mining reservoir dam. Given these reasons, it seems that Deadwater Slough could be a candidate for management or restoration actions to benefit local Chinook Salmon populations.

@@

```{r fall-bio-fig, fig.cap = "First bioenergetics plot."}
fall_bio_p = fall_only_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Scumulative fish eaten (g)",
                     breaks = seq(0, 200, by = 10),
                     sec.axis = sec_axis(~. / 12, 
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 20, by = 1))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)")

fall_bio_p

```{r full-yr-ng-fig, fig.cap = "Next bioenergetics plot."}
full_yr_ng_p = full_yr_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 30, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

full_yr_ng_p

```{r full-yr-10p-fig, eval = F, fig.cap = "Another bioenergetics plot."}
full_yr_10p_p = full_yr_10p_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 40, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

full_yr_10p_p

```{r alt-mod-fig, eval = F, fig.cap = "Another bioenergetics plot, again."}
alt_mod_p = alternative_mod %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 30, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

alt_mod_p

@@

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()

@@

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())

>>>>

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

@@

```{r schnabel-estimator, eval = T}
# using Schnabel and Schumacher-Eschmeyer estimators
N_mods = cmr_df %>%
  # filter out spring 2021 (no recapture)
  filter(event_name != "Spring_2021") %>%
  group_by(event_name, species, date) %>%
  summarise(n = n(),
            m = sum(capture_type == "Recapture", na.rm = T),
            u = sum(capture_type == "Mark" | is.na(capture_type)),
            R = sum(fish_status == "Live", na.rm = T),
            .groups = "drop") %>%
  group_by(event_name, species) %>%
  mutate(M = lag(R - m),
         M = replace_na(M, 0),
         M = cumsum(M)) %>%
  nest() %>%
  crossing(model = c("Schnabel",
                     "SchumacherEschmeyer")) %>%
  mutate(mr_model = map2(data,
                         model,
                         .f = function(x, y) {
                           with(x,
                                mrClosed(n = n,
                                         m = m,
                                         R = R,
                                         method = y,
                                         chapman.mod = TRUE))
                         }),
         N = map_dbl(mr_model,
                     .f = summary),
         CI = map(mr_model,
                  .f = confint),
         Lci = map_dbl(CI,
                       .f = function(x) x[1]), 
         Uci = map_dbl(CI,
                       .f = function(x) x[2])) %>%
  ungroup() %>%
  select(-CI)

@@

All of our estimators rely on meeting several assumptions:
1. The population in closed (no immigration, emigration, births or deaths during the sampling period).
1. Marking a fish does not affect its chances of being caught
1. No loss of marks
1. No marks are missed or mistaken
1. All fish have the same chance of being caught in the second (and subsequent) sample(s)

We applied several abundance estimators, which fell into two broad categories: single census and multiple census. For the single census estimators, we treated the first week of sampling as the mark event, and the second week as the recapture event, pooling data within each of those weeks. The multiple census estimators treat each day as a survey, and use information about the total marked fish from all previous surveys to infer the total abundance.

@@

The Lincoln-Petersen estimator can be biased with small samples, so we also investigated the Chapman-modified Lincoln-Petersen estimator which is shown below.

$$
\hat{N} = \frac{(M + 1)(n + 1)}{(m + 1)} - 1
$$

The Schnabel estimator is shown below, where the $M$, $n$ and $m$ are indexed by the sampling occasion, $i$. In our example, the sampling occasions are defined as each day of sampling. This estimator does not have an associate standard error, but 95% confidence intervals can be calculated. The Schnabel estimator is essentially a weighted average of a series of Lincoln-Petersen estimators (with a Chapman modification).

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_i}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$

There is another estimator for this type of multiple census surveys, called the Schumacher-Eschmeyer estimator, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. It is shown below.

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$
We additionally calculated the proportional stock density (PSD) of Northern Pikeminnow with 300 mm total length (TL) for stock and 400 mm TL for quality.
$$
PSD_{i} = 100 * \frac{FQ_{i}}{FS_{i}}
$$
where $FQ_{i}$ is the number of fish $\ge$ quality-length for species $i$, and $FS_{i}$ is the number of fish $\ge$ stock-length for species $i$.

@@

```{r shoup-temp-data}
temp_df = read_tsv(here("analysis/data/raw_data/USGS_13307000_Salmon_River_Nr_Shoup.txt"),
                   comment = "#") %>%
  rename(discharge_cfs = `47768_00060`,
         discharge_cfs_cd = `47768_00060_cd`,
         gage_ht_ft = `47769_00065`,
         gage_ht_ft_cd = `47769_00065_cd`,
         water_temp_c = `47773_00010`,
         water_temp_c_cd = `47773_00010_cd`)

@@

```{r abund-est}
all_est = LP_mod %>%
  select(event_name, species, model, N:Uci) %>%
  bind_rows(N_mods %>%
             select(event_name, species, model, N:Uci)) %>%
  mutate(range = Uci - Lci,
         prop_range = range / N) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  mutate(model = fct_recode(model,
                            "Schumacher-Eschmeyer" = "SchumacherEschmeyer")) %>%
  arrange(event_name, model)
```

@@

During 2019, 660 northern pikeminnow were gastric lavaged. Of those fish, 603 had empty stomachs, 57 had stomach contents, 12 had fish parts, and of those 12 with fish parts 1 MWF and 2 shiners. During the 2020 season 805 northern pikeminnow were gastric lavaged with 613 empty stomachs, 188 with stomach contents and 23 with fish parts. Out of the 23 stomachs with fish parts present 10 were unknown, 6 red sided shiner, 4 suckers, 2 sculpin, and 1 chinook. In 2021, we had 105 stomach samples with 98 being non-fish and 7 having fish or fish parts.

@@

Over the fall time period a single northern pikeminnow is modeled to eat 61 g of fish. If we extrapolate this out to the potential effect the estimated population of ***xx,xxx***, then ***xxx,xxx*** g of fish are being eaten each fall during the migration of juvenile Chinook salmon having the potential to consume ***xxx** juvenile Chinook as a discrete population.

Over the point of a whole year, a single northern pikeminnow is modeled to eat 283 g of fish and 613 g of other food types.

@@

The majority of fish collected during this study received gastric lavage, including some non-predatory species. To validate the efficacy of this method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. Our results support previous findings that gastric lavage effectively removes stomach contents (@Lott2020). Additionally, we found that the fish captured in net-traps had a similar proportion of stomach content samples. of those, 1,214 (76.2%) were completely emply, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish remnants. We were able to identify juvenile Red Shiners, a Largescale Sucker, Sculpin, Mountain Whitefish, and one Chinook Salmon within stomach contents, but most samples were too digested to identify to species.

@@

Although not formally assessed in this study, avian predators including Great Blue Herons *Ardea herodias* and Bald Eagles *Haliaeetus leucocephalus* are another potential source of mortality for juvenile salmon in the Deadwater Slough. The Deadwater Slough is recognized as an important bird watching and nesting area due to the associated riparian and backwater habitats [(Deadwater Slough - Audubon Important Bird Areas)](https://www.audubon.org/important-bird-areas/deadwater-slough#). Several piscivorous bird species have been documented using the Deadwater Slough that include the Common *Mergus merganser* and Hooded *Lophodytes cucullatus* mergansers, the Great Blue Heron, the Double-crested Cormorant *Phalacrocorax auritus*, and the Belted Kingfisher *Megaceryle alcyon* [@eBird2021]. During the intial sampling event in 2019, a two-person crew walked the entire reach and surrounding and upstream areas scanning for passive integrated transponder (PIT) tags. During that informal survey, nine PIT tags were recovered near active bird nests or in an upstream anastomizing reach where herons and eagles are prevalent, suggesting that mortality may have been a result of avian predation. The PIT tag histories in [PTAGIS](https://www.ptagis.org/) indicate these tags were implanted into a combination of juvenile Chinook Salmon (3), Sockeye Salmon (3), and steelhead (3). Avian predation contributes a major component of the total mortality for yearling Chinook Salmon in some locations in the lower Snake River and Columbia River, particularly at hydroelectric dams and within reservoirs [@Evans2012; -@Evans2016]; however, we did not observe large colonies of piscivorous birds within the study area. Although there is documentation of individual Double-crested Cormorants [@eBird2021] at the Deadwater Slough, the site is not within their breeding range, rather, it is part of a migration corridor. Given the current avian species known to occupy Deadwater Slough, it is unlikely that avian predation on juvenile salmonids is comparable to elsewhere in the Columbia River basin with large piscivorous bird colonies. Nevertheless, we hypothesize that the reservoir-like conditions at the Deadwater Slough may increase the probability of avian predation on juvenile Chinook Salmon from the many piscivorous birds known to use the site. Future estimates of predation would benefit from consideration of the contribution of piscivorous avian predators.

@@

We estimated that consumption of juvenile Chinook Salmon by northern pikeminnow in the Deadwater Slough potentially reduces annual adult returns by ***xxx - x,xxx*** to upriver populations. Presumably, that reduction in adult returns impacts both the ESA-listed natural populations in the Upper Salmon River MPG plus the two hatchery populations in the Upper Salmon, Pahsimeroi and Sawtooth hatcheries, which provide for recreational fishing opportunities. Fisheries managers desire increased adult returns to both aide in the recovery of natural populations as well as to provide additional harvest in recreational fisheries to boost local economies. A reduction in predation mortality at Deadwater Slough on juvenile Chinook Salmon has potential to benefit multiple upriver natural and hatchery populations, in contrast to typical tributary habitat rehabilition actions which typically benefit a single population. Moreover, the deepened, slack water conditions that favor northern pikeminnow at Deadwater Slough are indirectly the result of manmade activities i.e., the failure of a manmade mining reservoir dam. Given these reasons, it seems that Deadwater Slough could be a candidate for management or restoration actions to benefit local Chinook Salmon populations.

@@

@@

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

@@

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
